---
title: "Group presentation ML_EPI"
format: html
editor: visual
author: 
    - Morten
    - Christian
    - Jonas
---

# Load packages

```{r}
#| echo: false
#| warning: false


library(lattice)
library(NHANES)
library(dplyr)
library(caret)
library(randomForest)
library(here)
library(riskCommunicator)
library(gtsummary)
library(rpart.plot)
library(pROC)
```
# Objective

To evaluate risk scores performance between different models on an external population .

# Data set

NHANES

Framingham study

# Prepare data

```{r}
#| warning: false


data(framingham)
data("NHANES")

framingham <- framingham %>%
  select(DIABETES, SEX, AGE, BMI, SYSBP, DIABP, CURSMOKE, educ, HEARTRTE, TOTCHOL) %>% 
    mutate(TOTCHOL = TOTCHOL / 38.67)

framingham <- na.omit(framingham)
NHANES <- unique(NHANES)
NHANES <- NHANES %>%
  select(Diabetes, Gender, Age, BMI, BPSysAve, BPDiaAve, SmokeNow, Education, Pulse, TotChol)
NHANES <- na.omit(NHANES)

names(framingham) <- names(NHANES)
```



```{r}
#| include: false


NHANES$Diabetes <- ifelse(NHANES$Diabetes == "Yes", 1, 0)
NHANES$Diabetes <- as.factor(NHANES$Diabetes)
framingham$Diabetes <- as.factor(framingham$Diabetes)
  #summary(framingham$Gender)
  #summary(NHANES$Gender)
NHANES$Gender <- ifelse(NHANES$Gender == "female", 1, 0)
NHANES$Gender <- as.factor(NHANES$Gender)
framingham$Gender <- as.factor(framingham$Gender)
rename(framingham, Female = Gender)
rename(NHANES, Female = Gender)

framingham$Gender <- ifelse(framingham$Gender == 1, 0, 1)
framingham$Gender <- as.factor(framingham$Gender)

  #summary(framingham$Age)
  #summary(NHANES$Age)
  #summary(framingham$BMI)
  #summary(NHANES$BMI)
  #summary(framingham$BPSysAve)
  #summary(NHANES$BPSysAve)
  #summary(framingham$BPDiaAve)
  #summary(NHANES$BPDiaAve)
NHANES$BPDiaAve <- ifelse(NHANES$BPDiaAve == 0, NA, NHANES$BPDiaAve)
  NHANES <- na.omit(NHANES)
  #summary(framingham$SmokeNow)
  #summary(NHANES$SmokeNow)
NHANES$SmokeNow <- ifelse(NHANES$SmokeNow == "Yes", 1, 0)
NHANES$SmokeNow <- as.factor(NHANES$SmokeNow)
framingham$SmokeNow <- as.factor(framingham$SmokeNow)
  #summary(framingham$Education)
  #summary(NHANES$Education)
NHANES$Education <- as.numeric(NHANES$Education)
NHANES$Education <- ifelse(NHANES$Education == 4, 3, NHANES$Education)
NHANES$Education <- ifelse(NHANES$Education == 5, 4, NHANES$Education)
  range(NHANES$Education)
NHANES$Education <- as.factor(NHANES$Education)
framingham$Education <- as.factor(framingham$Education)

evaluation_data <- framingham 

framingham$Diabetes <- as.factor(framingham$Diabetes)

```



# Descriptive

```{r}

NHANES2 <- NHANES %>% 
    mutate(data_source = "NHANES")

framingham2 <- framingham %>% 
    mutate(data_source = "Framingham")


data_descriptive <- rbind(NHANES2, framingham2)

data_descriptive %>%
  tbl_summary(by = "data_source")
```

# Set up: Partition data into training/testing

```{r partition}
set.seed(123)

training.data <- createDataPartition(NHANES$Diabetes, p = 0.7, list = F)
train.data <- NHANES[training.data, ]
test.data <- NHANES[-training.data, ]
```



# Model development on NHANES

## Reference model logistic regression

- Upsampling because of low prevalence of diabetes

```{r}
set.seed(123)

control.obj <- trainControl(method = "cv", number = 10, sampling = "up")

logit.drisk <- train(
  Diabetes ~ .,
  data = train.data,
  method = "glm",
  family = "binomial",
  preProcess = c("center", "scale"),
  trControl = control.obj
)

logit.drisk$results
confusionMatrix(logit.drisk)
coef(logit.drisk$finalModel)
```

## Random forrest development

Prevalence of diabetes of too low in Framingham dataset gives problems in random forrest

```{r randomforest}
feat.count <- c((ncol(train.data) - 1), (ncol(train.data) - 1) / 2, sqrt(ncol(train.data) - 1))
grid.rf <- expand.grid(mtry = feat.count)

control.obj <- trainControl(method = "cv", number = 10, sampling = "up")
# control.obj <- trainControl("cv", number = 5)
tree.num <- seq(100, 500, by = 200)

results.trees <- list()
for (ntree in tree.num) {
  set.seed(123)
  rf.drisk <- train(
    Diabetes ~ .,
    data = train.data,
    method = "rf",
    trControl = control.obj,
    metric = "Accuracy",
    tuneGrid = grid.rf,
    importance = TRUE,
    ntree = ntree
  )
  index <- toString(ntree)
  results.trees[[index]] <- rf.drisk$results
}

output.rf.drisk <- bind_rows(results.trees, .id = "ntrees")
best.tune <- output.rf.drisk[which.max(output.rf.drisk[, "Accuracy"]), ]
best.tune$mtry
results.trees
```




### Final Random forrest model

```{r}
mtry.grid <- expand.grid(.mtry = best.tune$mtry)

set.seed(123)
rf.drisk.bt <- train(
  Diabetes ~ .,
  data = train.data,
  method = "rf",
  trControl = control.obj,
  metric = "Accuracy",
  tuneGrid = mtry.grid,
  importance = TRUE,
  ntree = as.numeric(best.tune$ntrees)
)

confusionMatrix(rf.drisk.bt)
varImp(rf.drisk.bt)
varImpPlot(rf.drisk.bt$finalModel)
```

## Ridge model

```{r}
set.seed(123)

#Create grid to search lambda
lambda <- 10^seq(-3,3, length=1000)

control.obj<-trainControl(method="cv", number=10, sampling="up")

#Note replacing tuneLength with tuneGrid
ridge.model<-train(
         Diabetes ~., data=train.data,
         method="glmnet", 
         trControl=control.obj, 
         preProc=c("center", "scale"), 
         tuneGrid=expand.grid(alpha = 0, lambda=lambda)
)

ridge.model$bestTune
  
# Model coefficients
coef(ridge.model$finalModel, ridge.model$bestTune$lambda)

# Make predictions in test set

ridge.pred <- ridge.model %>% predict(test.data)

# Model prediction performance
confusionMatrix(ridge.model)
```


# Evaluate models on test set in NHANES

### ROC curves

```{r}
#| echo: false

# Predict in test-set and output probabilities
rf.probs <- predict(rf.drisk.bt, test.data, type = "prob")

# Pull out predicted probabilities for Diabetes=Yes
rf.pp <- rf.probs[, 2]

# Predict in test-set using response type
logit.probs <- predict(logit.drisk, test.data, type = "prob")
logit.pp <- logit.probs[, 2]

# Predict in test-set using response type
ridge.probs <- predict(ridge.model, test.data, type = "prob")
ridge.pp <- ridge.probs[, 2]
```

```{r}
# Compute ROC curves
roc_logit <- roc(test.data$Diabetes, logit.pp, levels = rev(levels(test.data$Diabetes)))
roc_rf <- roc(test.data$Diabetes, rf.pp, levels = rev(levels(test.data$Diabetes)))
roc_ridge <- roc(test.data$Diabetes, ridge.pp, levels = rev(levels(test.data$Diabetes)))


# Plot ROC curves
plot(roc_logit, col = "blue", main = "ROC Curves for Logistic Regression and Random Forest")
lines(roc_rf, col = "red")
lines(roc_ridge, col = "green")
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "Ridge"), col = c("blue", "red","green"), lwd = 2)

```

### Calibration plot

```{r}
pred.prob <- data.frame(Class = test.data$Diabetes, logit = logit.pp, rf = rf.pp, ridge = ridge.pp)

calplot <- (calibration(Class ~ logit + rf + ridge, data = pred.prob, class = "1", cuts = 10))

xyplot(calplot, auto.key = list(columns = 2))
```

# Evaluate performance on external data from Framingham study

```{r}
#| echo: false

# Predict in test-set and output probabilities
rf.probs <- predict(rf.drisk.bt, evaluation_data, type = "prob")

# Pull out predicted probabilities for Diabetes=Yes
rf.pp <- rf.probs[, 2]

# Predict in test-set using response type
logit.probs <- predict(logit.drisk, evaluation_data, type = "prob")

logit.pp <- logit.probs[, 2]

# Predict in test-set using response type
ridge.probs <- predict(ridge.model, evaluation_data, type = "prob")
ridge.pp <- ridge.probs[, 2]
```

##  External performance on the Framingham study

### ROC curves

```{r}
# Compute ROC curves
roc_logit <- roc(evaluation_data$Diabetes, logit.pp, levels = rev(levels(evaluation_data$Diabetes)))
roc_rf <- roc(evaluation_data$Diabetes, rf.pp, levels = rev(levels(evaluation_data$Diabetes)))
roc_ridge <- roc(evaluation_data$Diabetes, ridge.pp, levels = rev(levels(evaluation_data$Diabetes)))


# Plot ROC curves
plot(roc_logit, col = "blue", main = "ROC Curves for Logistic Regression and Random Forest")
lines(roc_rf, col = "red")
lines(roc_ridge, col = "green")
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "Ridge"), col = c("blue", "red","green"), lwd = 2)
```

### Calibration plot

```{r}
pred.prob <- data.frame(Class = evaluation_data$Diabetes, logit = logit.pp, rf = rf.pp, ridge = ridge.pp)

calplot <- (calibration(Class ~ logit + rf + ridge, data = pred.prob, class = "1", cuts = 10))

xyplot(calplot, auto.key = list(columns = 2))
```


# Variable importance

```{r}
varImp(rf.drisk.bt)
varImpPlot(rf.drisk.bt$finalModel)
```

## Visualise the pathway of risk decision
```{r}
set.seed(123)
train.control <- trainControl(method = "cv", number = 10)

grid.singletree <- expand.grid(cp = seq(0.0005, 0.02, by = 0.001))

tree.diabetes <- train(
  Diabetes ~ .,
  data = train.data,
  method = "rpart",
  trControl = train.control,
  tuneGrid = grid.singletree
)


rpart.plot(tree.diabetes$finalModel)
```

# Ethical considerations

If by using a few easily obtainable features you can predict a person’s risk of developing diabetes, if that model was adopted by an insurance company, it might impact one's chance of getting an insurance and/or the price of the insurance

