---
title: "Exercise 1"
author: "JAS"
date: " "
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions for Exercise 1

R has a built-in dataset from the US National Health And Nutrition Examination Study (NHANES).
Using that dataset, you will try to predict Diabetes using a subset of variables and logistic regression.

Implement the following steps:

1. Restrict the NHANES data to the list of 11 variables below. 
        "Age", "Race1", "Education", "HHIncome", "Weight", "Height", "Pulse", "Diabetes", "BMI", "PhysActive",         "Smoke100"
2. Remove any duplicate observations using the unique() function.
3. Remove any observations with missing data  using the na.omit() function.
4. Partition your data into training and testing using a 70/30 split.
5. Set up your control settings to run 5-fold cross-validation
6. Within the training data, fit a logistic regression model using the glm method within caret.
7. Examine the average accuracy metrics across folds using confusionMatrix() function.
8. Apply your final model to the test set and obtain final evaluation metrics using confusion.Matrix()

### Chunk 1
This code chunk loads packages and performs the initial data cleaning steps. This includes subsetting to only the needed columns, removing duplicates and removing individuals with missing data.

```{r data_prep}
library(lattice)
library(NHANES)
library(dplyr)
library(caret)

data ("NHANES")

data.NHANES<-NHANES

#Remove duplicates
NHANES.unique<-unique(data.NHANES)

keep.var <- names(NHANES.unique) %in% c("Age", "Race1", "Education", "HHIncome", "Weight", "Height", "Pulse", "Diabetes", "BMI", "PhysActive", "Smoke100")

NHANES.subset<-NHANES.unique[keep.var]

#Examine data
str(NHANES.subset)
summary(NHANES.subset)



#Remove missings
NHANES.subset<-na.omit(NHANES.subset)

```

### Chunk 2

This chunk will partition data into training and testing using a 70/30 split

```{r partition}

#Remember to set seed for random actions
set.seed(123)
training.data <- NHANES.subset$Diabetes %>% createDataPartition(p=0.7, list=F)

#OR
#training.data<-createDataPartition(NHANES.subset$Diabetes, p=0.7, list=FALSE)

train.data<-NHANES.subset[training.data, ]
test.data<-NHANES.subset[-training.data, ]
```

### Chunk 3

This chunk will set up the training procedure in caret. This includes setting up 5-fold cross validation and constructing the logistic regression model.


```{r training}
set.seed(123)

train.control.settings<-trainControl(method="cv", 
                                     number=5)
logit.caret<-train(Diabetes ~ .,
                   data=train.data, 
                   method="glm", 
                   family="binomial", 
                   trControl=train.control.settings)

logit.caret$results
confusionMatrix(logit.caret)

```

### Chunk 4

This chunk applies the "best" model within the test set and obtains a final confusion matrix.

```{r evaluation}
set.seed(123)
pred.nhanes.test<-predict(logit.caret, test.data)

eval.results.4<-confusionMatrix(pred.nhanes.test, test.data$Diabetes, positive = "Yes")
print(eval.results.4)

#Examine coefficients for metric of variable importance
coef(logit.caret$finalModel)
```

```{r training_sampling}
set.seed(123)

train.control.settings<-trainControl(method="cv", 
                                     number=5, sampling="up")
logit.caret<-train(Diabetes ~ .,
                   data=train.data, 
                   method="glm", 
                   family="binomial", 
                   trControl=train.control.settings)

logit.caret$results
confusionMatrix(logit.caret)

```
